T-reX\@: Quantifying waste and material footprints in current and future \\ Life Cycle Assessment (LCA) databases

### 2024-05-02 latex_files/introduction.tex
commit text changes

\subsection{Background}\label{sec:intro-background}

The development of a `circular economy' has become a central focus in the frantic pursuit of sustainability objectives and the imperative curtailment of our environmental footprint within planetary boundaries~\citep{eu2019greendeal, eu2020circ,nl2023ceplan,nl2016ceplan,pardo2018ce,ellenmacarthur2015ce}. Fundamental to this development is a decrease in primary material consumption and a reduction of life cycle waste through the implementation of `re-X' strategies (e.g., refuse, rethink, design for---and implementation of---repair, remanufacturing and recycling)~\citep{reike2018rex, eu2022ecodesign, eu2022repair, eu2015reman}. In addition to circular economy goals, contemporary geo-political tensions in an ever more globalised economy have highlighted the vulnerability of many advanced economies to intentional supply disruptions---wrought as an act of competition or even outright hostility~\citep{jrc2023supplychain,hartley2024cepolitics,berry2023crm}.

The concept of the `footprint' as an environmental sustainability indicator began with the Ecological Footprint (EF)~\citep{wackernagel1994ecologicalfootprint} and after being popularised by the Carbon Footprint (CF)~\citep{cucek2015environmentalfootprints} the `footprint family' has adopted many additional metrics---albeit without yet coalescing into a coherent or consistent framework~\citep{giampietro2014footprintstonowhere, vanham2019footprints,ridoutt2013footprints}. More recently, the footprint collection has been extended to include the Material Footprint (MF)~\citep{weidmann2013materialfootprint}, which is often encountered in the `macro methods' of Industrial Ecology (IE), such as environmentally Extended Input-Output Analysis (EEIOA)~\citep{lenzen2022materialfootprint} and Material Flow Analysis (MFA)~\citep{schaffartzik2013mfafootprint}. Whether at the level of products, entire industries, nations or even continents, the material footprint aims to quantify the total material consumed in supply chains. It has been shown that the MF can be `highly representative of damage to human health and biodiversity'~\citep{steinmann2017resourcefootprints} and indeed, this metric was recently beatified by the United Nations (UN), becoming the `core official indicator' for targets 8.4 and 12.2 of the Sustainable Development Goals (SDGs)~\citep{lenzen2022materialfootprint}.

Though steadily emerging~\citep{laurenti2016wastefootprint,demirer2019wastefootprint,guillotreau2023wastefootprint}, the `Waste Footprint' (WF) metric, in contrast, has yet to enjoy such noble recognition for its attempts to measure and classify the waste generated by human activities. While it seems obvious that reducing life cycle waste is critical to the development of the circular economy~\citep{towa2020wastefootprint,ellenmacarthur2015ce}, the WF remains largely overlooked---especially in LCA models where waste itself is seldom apportioned any inherent environmental significance aside from the emissions related to its treatment~\citep{laurenti2023wastefootprint}.
Such neglect strikes the authors as unjustified, given that it has been repeatedly demonstrated that WFs have a strong association with environmental damage~\citep{laurenti2023wastefootprint,doka2024publications, ridoutt2010wasteimpacts,jaio2013wasteabsorbtionfootprint}. Additionally, it has been shown that the most vulnerable communities suffer disproportionately from the social and ecological impacts created by wasteful behaviours and sub-standard end-of-life treatment ~\citep{pellow2023envjusticewaste,akese2018envjustice}.

@@

Life Cycle Assessment (LCA) is a useful method for the holistic estimation of the environmental impacts of products and processes. LCA can comprehensively evaluate impacts across the entire life cycle---from `cradle to grave'--- often identifying critical hotspots and guiding prioritisation of actions~\citep{guinee2011lca}. The standard approach is to apply Life Cycle Impact Assessment (LCIA) methods (such as ReCiPe~\citep{huijbregts2016recipe} and CML~\citep{guinee2002cml}), which convert the inventory data into a set of impact scores based on the sum of the elementary flows (those between the technosphere and biosphere). These scores can then be aggregated into a single score for each impact category, that can be compared across products and processes. Extending on standard LCA is \textit{ex-ante} LCA, which employs future scenarios to construct `prospective background databases' to predict the impacts of supply chains that have not yet been (and may never be) realised~\citep{cucurachi2018exante,blanco2020exante}.

@@

\subsubsection{Waste in LCA}\label{sec:intro-waste}

@@

Existing LCA methodologies offer limited direct indicators at the impact assessment level, providing sparse information on the impacts of waste. This limitation becomes particularly evident when attempting to identify waste generation hotspots within a product's life cycle. Addressing these hotspots is crucial for advancing `circularity', but there is a lack of convenient and flexible methods to calculate waste flows in LCA and a pressing need for more comprehensive methods that can effectively quantify waste flows and, therefore, contribute to a greater understanding of a product's total environmental footprint.

\cite{laurenti2023wastefootprint} developed a method to calculate the waste footprint of a product or service based on solving the demand vectors of the activities, while also proposing simple measures to quantify waste `hazardousness' and `circularity' with this data. The method presented, however, is limited in its scope and flexibility, is computationally intensive, difficult to use, and not easily reproducible. As the authors acknowledge, due to the way that exchanges are identified and `tagged' this method also suffers from errors due to double counting.  In T-reX, the source database is deconstructed into a list of separate exchanges, and only those explicitly defined as hazardous are marked as such. The `T-reX' Python software tool presented herein provides a more flexible, transparent, and user-friendly approach to quantifying waste flows in LCA\@. Moreover, the utility of T-reX is not limited to waste, it can be used to categorise and aggregate any technosphere exchange of interest (or customised grouping thereof), such as water, gas, and critical raw materials.

@@

In the context of geo-political tension and a mineral-hungry `renewable energy transition', ever greater attention is being given to the security of material supply---especially for those considered `critical raw materials' (CRMs)~\citep{eu2023crmstudy,hool2023crm,mancini2013supplysecurity,jrc2023supplychain,hartley2024cepolitics,salviulo2021supplychain,iea2023crm,iea2023energytechperspectives}. While LCA seeks to model the technosphere (a.k.a.\ the anthroposphere) and its exchanges with the biosphere, its focus is often on the environmental impacts of the system---the endpoints---rather than the primary material flows themselves.

@@

Integration with the \texttt{premise} Python package~\citep{sacchi2022premise}---which connects the projections of integrated assessment models (IAMs) with current LCA databases---enables the user to create and manipulate prospective LCA databases. Frustratingly, the current utility of prospective databases---in general, and in particular for the waste sector---is constrained by the fact that, to date, the sectoral coverage of future life cycle inventories (LCIs) is largely confined to energy, steel, cement, and transport~\citep{sacchi2023premisedocs}. Indeed, despite the ever more critical need to reliably model and future waste management systems,~\cite{bisinella2024wastelca} reports an alarming lack of coherent development in this field.

>>>>

Development of a `circular economy' has become a central focus in the frantic pursuit of sustainability objectives toward curtailment of our environmental footprint within planetary boundaries~\citep{eu2019greendeal, eu2020circ,nl2023ceplan,nl2016ceplan,pardo2018ce,ellenmacarthur2015ce}. Fundamental to this development is a decrease in primary material consumption and a reduction of life cycle waste through the implementation of `re-X' strategies (e.g., refuse, rethink, design for---and implementation of---repair, remanufacturing and recycling)~\citep{reike2018rex, eu2022ecodesign, eu2022repair, eu2015reman}. In addition to circular economy goals, contemporary geopolitical tensions in an ever more globalised economy have highlighted the vulnerability of many advanced economies to intentional supply disruptions---wrought as an act of competition or even outright hostility~\citep{jrc2023supplychain,hartley2024cepolitics,berry2023crm}.

The concept of the `footprint' as an environmental sustainability indicator began with the Ecological Footprint (EF)~\citep{wackernagel1994ecologicalfootprint} and after being popularised by the Carbon Footprint (CF)~\citep{cucek2015environmentalfootprints} the `footprint family' has adopted many additional metrics---albeit without yet coalescing into a coherent or consistent framework~\citep{giampietro2014footprintstonowhere, vanham2019footprints,ridoutt2013footprints}. More recently, the footprint collection has been extended to include the Material Footprint (MF)~\citep{weidmann2013materialfootprint}, which is often encountered in the `macro methods' of Industrial Ecology (IE), such as environmentally Extended Input-Output Analysis (EEIOA)~\citep{lenzen2022materialfootprint} and Material Flow Analysis (MFA)~\citep{schaffartzik2013mfafootprint}. Whether at the level of products, entire industries, nations, or even continents, the material footprint aims to quantify the total material consumed in supply chains. It has been shown that the MF can be `highly representative of damage to human health and biodiversity'~\citep{steinmann2017resourcefootprints} and indeed, this metric was recently beatified by the United Nations (UN), becoming the `core official indicator' for targets 8.4 and 12.2 of the Sustainable Development Goals (SDGs)~\citep{lenzen2022materialfootprint}.

Though steadily emerging~\citep{laurenti2016wastefootprint,demirer2019wastefootprint,guillotreau2023wastefootprint}, the `Waste Footprint' (WF) metric is less well-developed. While it seems obvious that reducing life cycle waste is critical to the development of the circular economy~\citep{towa2020wastefootprint,ellenmacarthur2015ce}, the WF remains largely overlooked---especially in LCA models where waste itself is seldom apportioned any inherent environmental significance aside from the emissions related to its treatment~\citep{laurenti2023wastefootprint}. Such neglect strikes the authors as unjustified, given that it has been repeatedly demonstrated that WFs have a strong association with environmental damage~\citep{laurenti2023wastefootprint,doka2024publications, ridoutt2010wasteimpacts,jaio2013wasteabsorbtionfootprint}. Additionally, it has been shown that the most vulnerable communities suffer disproportionately from the social and ecological impacts created by wasteful behaviours and substandard end-of-life treatment~\citep{pellow2023envjusticewaste,akese2018envjustice}.

@@

Life Cycle Assessment (LCA) is a useful method for the holistic estimation of the environmental impacts of products and processes. LCA can comprehensively evaluate impacts across the entire life cycle---from `cradle to grave'--- often identifying critical hotspots and guiding prioritisation of action~\citep{guinee2011lca}. The standard approach is to apply Life Cycle Impact Assessment (LCIA) methods (such as ReCiPe~\citep{huijbregts2016recipe} and CML~\citep{guinee2002cml}), which convert the inventory data into a set of impact scores based on the sum of the elementary flows (those between the technosphere and biosphere). These scores can then be aggregated into a single score for each impact category, that can be compared across products and processes. Extending on standard LCA is \textit{ex-ante} LCA, which employs future scenarios to construct `prospective background databases' to predict the impacts of supply chains that have not yet been (and may never be) realised~\citep{cucurachi2018exante,blanco2020exante}.

@@

Integration with the \texttt{premise} Python package~\citep{sacchi2022premise}---which connects the projections of integrated assessment models (IAMs) with current LCA databases---enables the user to create and manipulate prospective LCA databases. Frustratingly, the current utility of prospective databases---in general, and in particular for the waste sector---is constrained by the fact that, to date, the sectoral coverage of future life cycle inventories (LCIs) is largely confined to energy, steel, cement, and transport~\citep{sacchi2023premisedocs}. Despite the ever more critical need to reliably model and future waste management systems,~\cite{bisinella2024wastelca} reports an alarming lack of coherent development in this field.@\texttt{Premise} offers an excellent structure for the integration of future waste and material flow modelling in LCA databases, but its potential contribution to the development of the circular economy cannot be realised if the scientific resources for data collection and validation are not forthcoming. The authors believe that there is an urgent need for research investment in this area.

\subsection{Background and need}\label{sec:intro-background}
\subsubsection{Waste in LCA}\label{sec:intro-waste}

@@

Some existing LCA methodologies offer direct indicators at the impact assessment level, but generally, there is little attention given to the impacts of waste in LCA\citep{lauran2020abioticdepletion}. This limitation becomes particularly evident when attempting to identify waste generation hotspots within a product's life cycle. Addressing these hotspots is crucial for advancing `circularity', but there is a lack of convenient and flexible methods to calculate waste flows in LCA and a pressing need for more comprehensive methods that can effectively quantify waste flows and, therefore, contribute to a greater understanding of a product's total environmental footprint.

\cite{laurenti2023wastefootprint} developed a method to calculate the waste footprint of a product or service based on solving the demand vectors of the activities, while also proposing simple measures to quantify waste `hazardousness' and `circularity' with this data. The method presented, as the authors explain, suffers from errors due to double counting due to the way that exchanges are identified and `tagged' . In T-reX, the source database is deconstructed into a list of separate exchanges, and only those explicitly defined as hazardous are marked as such. The `T-reX' Python software tool presented herein provides a more flexible, transparent, and user-friendly approach to quantifying waste flows in LCA\@. Moreover, the utility of T-reX is not limited to waste, it can be used to categorise and aggregate any technosphere exchange of interest (or customised grouping thereof), such as water, gas, and critical raw materials.

@@

In the context of geopolitical tension and a mineral-hungry `renewable energy transition', greater attention is being given to the security of material supply---especially for those considered `critical raw materials' (CRMs)~\citep{eu2023crmstudy,hool2023crm,mancini2013supplysecurity,jrc2023supplychain,hartley2024cepolitics,salviulo2021supplychain,iea2023crm,iea2023energytechperspectives}. While LCA seeks to model the technosphere (a.k.a.\ the anthroposphere) and its exchanges with the biosphere, its focus is often on the environmental impacts of the system rather than the primary material flows themselves.

@@

### 2024-05-02 latex_files/main.tex
update-pdf
commit text changes

\begin{document}
\vspace{20m}
\includegraphics[width=10cm]{graphical-abstract.pdf}

@@

    \title{T-reX: Quantifying waste and material footprints in current and future Life Cycle Assessment (LCA) databases}
    \author[1]{Stewart Charles McDowall\corref{cor1}}
    \ead{s.c.mcdowall@cml.leidenuniv.nl}
    \author[1]{Elizabeth Lanphear}
    \author[1]{Stefano Cucurachi}
    \author[1]{Carlos Felipe Blanco}

@@

        We introduce T-reX, a Python tool extending the brightway ecosystem, to allow facile quantification of user-defined supply chain demands in current and prospective scenarios. T-reX streamlines database manipulation for LCA practitioners and integrates methods to aggregate and analyse waste and material flows, facilitating rapid hotspot identification.

        A case study of five lithium-ion batteries demonstrates T-reX's utility, quantifying categorised waste and material inventory footprints and their potential environmental burdens. T-reX can aid sustainable decision-making and contribute to the development of `circular economy' by facilitating analysis of material consumption and waste generation in LCA.

@@

% Back matter
\section*{Data availability}
All data used is publically available online under the noted sources. T-reX is fully open source and dedicated to the public domain under the CC-0 licence. It installable via the Python Package Index (PyPI) under the name `T\_reX\_LCA' \url{https://pypi.org/project/T_reX_LCA}.
The full source code for T-reX is archived by Zenodo at \url{https://zenodo.org/records/10925359} and the development server is hosted by Github at \url{https://www.github.com/Stew-McD/T-reX}. A user guide with comprehensive documentation are presented at \url{https://T-reX.readthedocs.io}.

>>>>

\begin{document}
\vspace{20em}
\includegraphics[width=10cm]{graphical-abstract.pdf}

@@

    \title{T-reX\@: Quantifying waste and material footprints in current and future \\ Life Cycle Assessment (LCA) databases}
    \author[1]{Stewart Charles McDowall\corref{cor1}}
    \ead{s.c.mcdowall@cml.leidenuniv.nl}
    \author[1]{Elizabeth Lanphear}
    \author[1]{Stefano Cucurachi}
    \author[1]{Carlos Felipe Blanco}

@@

        We introduce T-reX, a Python tool extending the brightway ecosystem to allow facile quantification of user-defined supply chain demands in current and prospective scenarios. T-reX streamlines database manipulation for LCA practitioners and integrates methods to aggregate and analyse waste and material flows, facilitating rapid hotspot identification.

        A case study of five lithium-ion batteries demonstrates T-reX's utility, quantifying categorised waste and material inventory footprints and their potential environmental burdens. T-reX can aid sustainable decision-making and contribute to the development of a `circular economy' by facilitating analysis of material consumption and waste generation in LCA.

@@

% Back matter
\section*{Data availability}
All data used is publically available online under the noted sources. T-reX is fully open source and dedicated to the public domain under the CC-0 licence. It is installable via the Python Package Index (PyPI) under the name `T\_reX\_LCA' \url{https://pypi.org/project/T_reX_LCA}.
The full source code for T-reX is archived by Zenodo at \url{https://zenodo.org/records/10925359} and the development server is hosted by Github at \url{https://www.github.com/Stew-McD/T-reX}. A user guide with comprehensive documentation are presented at \url{https://T-reX.readthedocs.io}.

### 2024-05-02 latex_files/methodology.tex
accept LL changes
commit text changes

This section is divided into two parts. In \autoref{sec:method-TreX}, we describe the T-reX program, and in \autoref{sec:method-casestudy}, we describe the methodology used to calculate the waste and material inventory footprints in a case study of five Li-ion batteries.

@@

T-reX can be used directly from the command line, or imported as a Python package where the user can access the individual functions and modules. In the simplest case, the user can run the program with the default settings, which will calculate the waste and material footprint of the \texttt{ecoinvent}database. The user can adapt the settings of T-reX as desired, to calculate alternative or additional waste and material inventory footprints, use a custom database, or a `prospective' database based on future scenarios by implementing the premise \texttt{premise} with one of T-reX's internal modules.

@@

In addition to the waste categories, T-reX can incorporate any number of material demand categories, the defaults being based on the EU Critical Raw Materials (CRM) list for 2023~\citep{eu2023crmstudy}, which contains 30 materials considered essential to the EU economy and also at risk of supply disruption. Especially valuable in the case of CRMs is T-reX's ability to uncover and quantify these threads of background consumption, which can be inconspicuously embedded in products or consumed in seemingly distant supply chain activities. The identity of the materials considered and their categorical groupings are easily customisable by the user, and in further materials of interest to the authors were added to the example Li-ion case study, including helium, electricity, petroleum, sand, water, and natural gas. The full list of 59 materials included in the default configuration is provided in the supplementary material in ~\autoref{sec:supplementary}.

@@

% As presented in~\autoref{sec:intro-background},~\autoref{sec:intro-waste}, and~\autoref{sec:intro-material} of the introduction, there are some existing material demand methods in the standard LCIA method sets, including the CSI (which provides only an aggregated, abstracted endpoint in units of kg~Si~eq)~\citep{arvidsson2020csi} and the (deprecated) EDIP 2003 material use indicators (which provide endpoints in fundamental units)~\citep{hauschild2003edip}.

@@

\begin{figure}[H]
    \centering
    \caption{Workflow of T-reX. Application of T-reX to one or more LCA databases creates a \texttt{brightway} project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}
    \includegraphics[width=14cm]{figures/T-reX_method.pdf}\label{fig:methods-flowchart}
\end{figure}

@@

We investigated five types of Li-ion batteries, each represented by their unamended global market activities:
\begin{itemize}[itemsep=0pt]
    \item Li-ion, LFP, rechargeable, prismatic
    \item Li-ion, LiMn\(_2\)O\(_4\), rechargeable, prismatic
    \item Li-ion, NMC111, rechargeable, prismatic
    \item Li-ion, NMC811, rechargeable, prismatic
    \item Li-ion, NCA, rechargeable, prismatic
\end{itemize}

@@

\begin{itemize}[itemsep=0pt]
    \item ReCiPe 2016 v1.03, midpoint (I)
    \item EF v3.0 no LT
    \item EDIP 2003 no LT
    \item CSI 2020
\end{itemize}

The source of the life cycle inventory data for this case study was \texttt{ecoinvent}, version 3.9.1, with the `cutoff' attributional system model. Additionally, T-reX was used to create prospective database sets using the functionality of \texttt{premise} with the REMIND model and the baseline scenario SSP2 `middle of the road' with the following Representative Concentration Pathways (RCPs):
\begin{itemize}
    \item SSP2-base: representing an approximate 3.5°C increase in global temperatures to 2100
    \item SSP2-PkBudg500: representing the achievement of Paris climate goals, ca. 1.3°C increase to 2100
\end{itemize}

>>>>

This section is divided into two parts. In \autoref{sec:method-TreX}, we describe the T-reX program, and in \autoref{sec:method-casestudy}, we describe the methodology used to calculate the waste and material inventory footprints in a case study of five Li-ion batteries, highlighting inconsistencies in LCA methodology and probing departures from projections in prospective models.

@@

T-reX can be used directly from the command line, or imported as a Python package with which the user can access the individual functions and modules. In the simplest case, the user can run the program with the default settings, which will calculate the waste and material footprint of the \texttt{ecoinvent}database. The user can adapt the settings of T-reX as desired, to calculate alternative or additional waste and material inventory footprints, use a custom database, or a `prospective' database based on future scenarios by implementing the premise \texttt{premise} with one of T-reX's internal modules.

@@

In addition to the waste categories, T-reX can incorporate any number of material demand categories, the defaults being based on the EU Critical Raw Materials (CRM) list for 2023~\citep{eu2023crmstudy}, which contains 30 materials considered essential to the EU economy and also at risk of supply disruption. Especially valuable in the case of CRMs is T-reX's ability to uncover and quantify these threads of background consumption, which can be inconspicuously embedded in products or consumed in seemingly distant supply chain activities. The identity of the materials considered, and their categorical groupings, are easily customisable by the user. The full list of 59 materials included in the default configuration is provided in the supplementary material in~\autoref{sec:supplementary}.

@@

@@

\begin{figure}[H]
    \centering
    \caption{The workflow of T-reX presented in a generalised (a) and computational (b) format. Application of T-reX to one or more LCA databases creates a \texttt{brightway} project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}
    \includegraphics[width=9cm]{figures/T-reX_method.pdf}\label{fig:methods-flowchart}
\end{figure}

@@

We investigated five types of Li-ion batteries in the \texttt{ecoinvent} 3.9.1 `cutoff' LCA database~\citep{ecoinvent2016version3}, each represented by their unamended global market inventories:
\begin{itemize}[itemsep=0pt]
    \item Li-ion, LFP, rechargeable, prismatic
    \item Li-ion, LiMn\(_2\)O\(_4\), rechargeable, prismatic
    \item Li-ion, NMC111, rechargeable, prismatic
    \item Li-ion, NMC811, rechargeable, prismatic
    \item Li-ion, NCA, rechargeable, prismatic
\end{itemize}

@@

\begin{itemize}[itemsep=0pt]
    \item ReCiPe 2016 v1.03, midpoint (I)~\citep{huijbregts2016recipe}
    \item EF v3.0 no LT~\citep{eu2023ef}
    \item EDIP 2003 no LT~\citep{hauschild2003edip}
    \item CSI 2020~\citep{arvidsson2020csi}
\end{itemize}

Additional to the present-day databases, T-reX created prospective database sets based on \texttt{ecoinvent} 3.9.1 using the functionality of \texttt{premise} with the REMIND model and the baseline scenario SSP2 `middle of the road' with the following Representative Concentration Pathways (RCPs):
\begin{itemize}
    \item SSP2-base: representing an approximate 3.5°C increase in global temperatures to 2100
    \item SSP2-PkBudg500: representing the achievement of Paris climate goals, ca. 1.3°C increase to 2100
\end{itemize}
